---
title: "Swire Capstone EDA"
author: "Nidal Arain"
date: "2025-02-03"
output: 
  html_document: 
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## loading data

```{r}
library(dplyr)
library(tidyverse)
library(readxl)

transactional_data <- read.csv("transactional_data.csv")
customer_profile_data <- read.csv("customer_profile.csv")
customer_address_zip_mapping_data <- read.csv("customer_address_and_zip_mapping.csv")
delivery_cost_data <- read_excel("delivery_cost_data.xlsx")
```

## Transactional Data EDA

```{r}
# Convert Date Format
transactional_data <- transactional_data %>%
  mutate(TRANSACTION_DATE = as.Date(TRANSACTION_DATE, format = "%m/%d/%Y"))

# Overview of the dataset structure
str(transactional_data)
# Get summary statistics for each column
summary(transactional_data)

```


# Find High-Growth Potential Customers
identify customers who are just below the threshold (e.g., 300–399 gallons) and see if they are growing.

These customers might benefit from active sales support to push them over 400 gallons.

```{r}
growth_potential <- customer_volume %>%
  filter(total_gallons >= 300 & total_gallons < 400) %>%
  arrange(desc(total_gallons))

print(growth_potential)

```


# Investigate Order Type & Growth

```{r}
order_type_growth <- transactional_data %>%
  group_by(ORDER_TYPE) %>%
  summarise(
    total_cases = sum(ORDERED_CASES, na.rm = TRUE),
    total_gallons = sum(ORDERED_GALLONS, na.rm = TRUE),
    total_volume = total_cases + total_gallons
  ) %>%
  arrange(desc(total_volume))

print(order_type_growth)


```

# Delivery Efficiency by Customer Group
```{r}
fulfillment_by_customer <- transactional_data %>%
  group_by(CUSTOMER_NUMBER) %>%
  summarise(
    total_loaded = sum(LOADED_CASES, na.rm = TRUE),
    total_delivered = sum(DELIVERED_CASES, na.rm = TRUE),
    fulfillment_rate = total_delivered / total_loaded
  ) %>%
  arrange(fulfillment_rate)

print(head(fulfillment_by_customer, 10))  # Check customers with lowest fulfillment

```

# Examine Numeric Distributions & Outliers
```{r}
negative_delivered <- transactional_data %>%
  filter(DELIVERED_CASES < 0 | DELIVERED_GALLONS < 0)

summary(negative_delivered)
```

The summary shows that there are a number of transactions with negative values for delivered cases and gallons. Here are some observations and follow-up questions:

Negative Delivered Values:
DELIVERED_CASES:
Min: -3132, 1st Quartile: -12, Median: -1, Mean: -35.3, Max: 1056
DELIVERED_GALLONS:
Min: -1792.5, 1st Quartile: -5, Median: 0, Mean: -6.725, Max: 387.5
Interpretation:

These negatives could be because of returns, adjustments, or corrections. The relatively low median (e.g., -1 for cases, 0 for gallons) suggests that many of these adjustments might be small, but there are some substantial negative outliers.


Next Steps/ Questions 
Do we have any business context on these negative values? Do they represent product returns, cancellations, or error corrections?


```{r}
negative_delivered %>%
  group_by(ORDER_TYPE) %>%
  summarise(count = n(), mean_delivered_cases = mean(DELIVERED_CASES, na.rm = TRUE))

```


# Distribution by Order Type
```{r}
transactional_data %>%
  group_by(ORDER_TYPE) %>%
  summarise(count = n()) %>%
  arrange(desc(count))


```

# Investigate Order Type Trends Over Time
```{r}
library(ggplot2)

order_type_trend <- transactional_data %>%
  group_by(YEAR, WEEK, ORDER_TYPE) %>%
  summarise(total_cases_ordered = sum(ORDERED_CASES, na.rm = TRUE),
            total_cases_delivered = sum(DELIVERED_CASES, na.rm = TRUE)) %>%
  ungroup()

ggplot(order_type_trend, aes(x = WEEK, y = total_cases_ordered, color = ORDER_TYPE)) +
  geom_line() +
  facet_wrap(~YEAR) +
  labs(title = "Weekly Ordered Cases by Order Type",
       x = "Week",
       y = "Total Ordered Cases",
       color = "Order Type")

```
Observations from the Weekly Ordered Cases by Order Type Chart

MYCOKE LEGACY Declining in 2024
In 2023, MYCOKE LEGACY (green line) had the highest number of ordered cases, peaking above 90,000 weekly.
In 2024, MYCOKE LEGACY orders decline sharply around week 30 and seem to be phased out.
This aligns with the business transition to MYCOKE360 (new digital platform).
MYCOKE360 Increasing in 2024

MYCOKE360 (cyan/blue-green line) shows an increasing trend in 2024.
There's a major spike around week 30, likely when many customers transitioned from MYCOKE LEGACY to MYCOKE360.
SALES REP Orders Are Stable but Show Slight Decline

SALES REP (pink line) remains relatively stable over both years, although it seems to decline slightly towards the end of 2024.
This could suggest that more customers are shifting to digital ordering methods (MYCOKE360, EDI, or CALL CENTER).
CALL CENTER & EDI Orders Are Consistent

CALL CENTER (red line) maintains a steady trend across both years.
EDI (brown/yellow line) fluctuates but remains fairly stable.
Null and OTHER Categories Need Investigation

The null category (purple line) is surprisingly large.
This needs further investigation:
Is this an issue with missing order types?
Are these manual adjustments, corrections, or unclassified orders?


## Customer Profile Data

```{r}

# Overview of dataset
str(customer_profile_data)

# Summary statistics
summary(customer_profile_data)

# Check for missing values
colSums(is.na(customer_profile_data))

```
```{r}
#converting to date type
customer_profile_data <- customer_profile_data %>%
  mutate(
    FIRST_DELIVERY_DATE = as.Date(FIRST_DELIVERY_DATE, format = "%m/%d/%Y"),
    ON_BOARDING_DATE = as.Date(ON_BOARDING_DATE, format = "%m/%d/%Y")
  )

```



#Compare Local Market Partners vs. All Customers

```{r}
local_vs_all <- customer_profile_data %>%
  group_by(LOCAL_MARKET_PARTNER) %>%
  summarise(
    num_customers = n(),
    avg_ordered_gallons = mean(transactional_data$ORDERED_GALLONS, na.rm = TRUE),
    avg_total_volume = mean(transactional_data$ORDERED_CASES + transactional_data$ORDERED_GALLONS, na.rm = TRUE)
  )

print(local_vs_all)

```

#Identify Customer Attributes Correlated with High-Volume Orders
```{r}
high_volume_customers <- transactional_data %>%
  group_by(CUSTOMER_NUMBER) %>%
  summarise(total_cases = sum(ORDERED_CASES, na.rm = TRUE),
            total_gallons = sum(ORDERED_GALLONS, na.rm = TRUE),
            total_volume = total_cases + total_gallons) %>%
  left_join(customer_profile_data, by = "CUSTOMER_NUMBER") %>%
  mutate(high_volume = ifelse(total_gallons >= 400, "Above 400 Gallons", "Below 400 Gallons"))

# Compare characteristics of high-volume vs. low-volume customers
high_volume_summary <- high_volume_customers %>%
  group_by(high_volume, LOCAL_MARKET_PARTNER, FREQUENT_ORDER_TYPE, TRADE_CHANNEL) %>%
  summarise(count = n()) %>%
  arrange(desc(count))

print(high_volume_summary)

```

# Identify Local Market Partners Near the 400-Gallon Threshold
```{r}
lmp_growth_potential <- high_volume_customers %>%
  filter(LOCAL_MARKET_PARTNER == TRUE & total_gallons >= 300 & total_gallons < 400) %>%
  arrange(desc(total_gallons))

print(lmp_growth_potential)

```







#Investigate Trade Channel & Sub-Trade Channel Impact
```{r}
trade_channel_summary <- customer_profile_data %>%
  group_by(TRADE_CHANNEL) %>%
  summarise(num_customers = n()) %>%
  arrange(desc(num_customers))

print(trade_channel_summary)

ggplot(trade_channel_summary, aes(x = reorder(TRADE_CHANNEL, num_customers), y = num_customers)) +
  geom_bar(stat = "identity", fill = "lightgreen") +
  coord_flip() +
  labs(title = "Distribution of Trade Channels",
       x = "Trade Channel",
       y = "Number of Customers")

```
1️⃣ Most Customers Belong to the Top 5 Trade Channels
FAST CASUAL DINING (6,016 customers) and COMPREHENSIVE DINING (4,765 customers) dominate the dataset.
GENERAL RETAILER (2,846) and OTHER DINING & BEVERAGE (2,797) are also significant.
OUTDOOR ACTIVITIES (2,194) is surprisingly high, suggesting seasonal or location-based demand.
Implication:

The majority of customers are in the food & beverage sector.
Fast Casual & Comprehensive Dining likely have different order patterns than Retailers or Recreation businesses.
Do high-volume customers belong to certain trade channels? → We should check.
2️⃣ Smallest Categories May Be Outliers
INDUSTRIAL, DEFENSE, TRAVEL, BULK TRADE, SUPERSTORE, and PHARMACY RETAILER have fewer than 200 customers each.
LARGE-SCALE RETAILER has only 1 customer.
Implication:

These sectors may not be relevant for ARTM strategy.
Should we remove them from modeling or treat them separately?
```{r}
trade_channel_volume <- transactional_data %>%
  left_join(customer_profile_data, by = "CUSTOMER_NUMBER") %>%
  group_by(TRADE_CHANNEL) %>%
  summarise(
    avg_total_volume = mean(ORDERED_CASES + ORDERED_GALLONS, na.rm = TRUE),
    num_customers = n()
  ) %>%
  arrange(desc(avg_total_volume))

print(trade_channel_volume)

```







## Customer Address & ZIP Mapping
Key EDA Goals for ZIP Mapping
1️⃣ Identify Geographic Trends in High-Volume vs. Low-Volume Customers

Are certain regions more likely to exceed the 400-gallon threshold?
Can we spot clusters of high-growth potential customers?
2️⃣ Analyze Local Market Partner Distribution by Region

Are LMPs concentrated in specific areas?
Could regional differences impact ARTM transition decisions?
3️⃣ Compare Delivery Efficiency by ZIP Code

Are some ZIP codes experiencing more fulfillment issues (low delivered-to-ordered ratio)?
Could logistics challenges make some areas better candidates for ARTM?


```{r}
# Overview of dataset
str(customer_address_zip_mapping_data)

# Summary statistics
summary(customer_address_zip_mapping_data)

# Check for missing values
colSums(is.na(customer_address_zip_mapping_data))

```

# Identify High-Volume vs. Low-Volume ZIP Codes
These low-volume ZIPs may be better candidates for ARTM since direct truck deliveries might not be cost-effective.
```{r}
# Rename 'zip' column in customer_address_zip_mapping_data to match the other datasets
customer_address_zip_mapping_data <- customer_address_zip_mapping_data %>%
  rename(ZIP_CODE = zip)

# Perform the join and fix grouping issue
zip_volume_analysis <- transactional_data %>%
  left_join(customer_profile_data, by = "CUSTOMER_NUMBER") %>%
  left_join(customer_address_zip_mapping_data, by = "ZIP_CODE") %>%
  group_by(ZIP_CODE) %>%  # Fix grouping function
  summarise(
    total_customers = n(),
    avg_total_gallons = mean(ORDERED_GALLONS, na.rm = TRUE),
    high_volume_customers = sum(ORDERED_GALLONS >= 400, na.rm = TRUE)
  ) %>%
  arrange(desc(high_volume_customers))

print(zip_volume_analysis)

# Visualize top ZIP codes with high-volume customers
ggplot(zip_volume_analysis %>% filter(high_volume_customers > 0), 
       aes(x = reorder(ZIP_CODE, high_volume_customers), y = high_volume_customers)) +
  geom_bar(stat = "identity", fill = "blue") +
  coord_flip() +
  labs(title = "Top ZIP Codes by High-Volume Customers (400+ Gallons)",
       x = "ZIP Code",
       y = "Number of High-Volume Customers")


```




#Analyze LMP Distribution by ZIP
```{r}
lmp_zip_distribution <- customer_profile_data %>%
  filter(LOCAL_MARKET_PARTNER == TRUE) %>%
  left_join(customer_address_zip_mapping_data, by = "ZIP_CODE") %>%
  group_by(ZIP_CODE) %>%
  summarise(lmp_count = n()) %>%
  arrange(desc(lmp_count))

print(lmp_zip_distribution)
```


#Compare LMP-Heavy ZIPs with High-Volume ZIPs
ZIPs with both high LMP density & low high-volume customers should likely be moved to ARTM to reduce delivery costs.

ZIPs with high high-volume customer density should likely stay on direct delivery (red trucks) to retain and grow these accounts.

ZIPs in between (moderate LMP density and moderate high-volume customers) may need a hybrid approach (case-by-case evaluation).
```{r}
# Merge the two datasets to compare LMP-heavy and high-volume ZIPs
zip_comparison <- lmp_zip_distribution %>%
  full_join(zip_volume_analysis, by = "ZIP_CODE") %>%
  replace_na(list(lmp_count = 0, high_volume_customers = 0)) %>%  # Fill NAs with 0
  mutate(lmp_density = lmp_count / total_customers,  # % of customers that are LMPs
         high_volume_density = high_volume_customers / total_customers) %>%
  arrange(desc(lmp_density))

print(zip_comparison)

# Visualize LMP-heavy vs. High-Volume ZIPs
ggplot(zip_comparison, aes(x = lmp_density, y = high_volume_density, label = ZIP_CODE)) +
  geom_point(color = "blue", alpha = 0.6) +
  geom_text(size = 3, vjust = 1.5, check_overlap = TRUE) +
  labs(title = "LMP Density vs. High-Volume Customer Density by ZIP",
       x = "Percentage of Customers That Are LMPs",
       y = "Percentage of Customers That Are High-Volume (400+ Gallons)") +
  theme_minimal()

```



## Delivery Cost Data
Cost Structure Breakdown

Median delivery cost ranges from $0.37 to $8.58 per unit.
The highest median delivery cost is $8.06 per case (likely for low-volume customers).
The costs decrease as volume increases, which aligns with economies of scale.
```{r}
# Overview of dataset
str(delivery_cost_data)

# Summary statistics
summary(delivery_cost_data)

# Check for missing values
colSums(is.na(delivery_cost_data))

```


#Analyze Cost Per Unit by Product Category

1️⃣ Bottles and Cans Are More Expensive to Deliver

Median Cost: $3.13 per case
Average Cost: $3.51 per case
Max Cost: $8.58 per case → Very high for low-volume deliveries.
2️⃣ Fountain Drinks Have Lower Delivery Costs

Median Cost: $1.47 per gallon
Average Cost: $1.68 per gallon
Max Cost: $4.62 per gallon → Still lower than Bottles and Cans.


🚀 Key Insights for ARTM Strategy
Low-volume Bottles and Cans deliveries are very expensive (up to $8.58 per case).
Customers below 400 gallons, especially those ordering Bottles and Cans, are strong ARTM candidates.

Fountain drink customers generally have lower costs.
They might justify staying on direct delivery if their volume is high enough.
SCCU can prioritize ARTM for customers ordering Bottles & Cans in small volumes.


```{r}
# Standardize column names by replacing spaces with underscores
colnames(delivery_cost_data) <- gsub(" ", "_", colnames(delivery_cost_data))

# Verify that column names are now correct
colnames(delivery_cost_data)
cost_by_category <- delivery_cost_data %>%
  group_by(Applicable_To) %>%
  summarise(
    median_cost = median(Median_Delivery_Cost, na.rm = TRUE),
    avg_cost = mean(Median_Delivery_Cost, na.rm = TRUE),
    min_cost = min(Median_Delivery_Cost, na.rm = TRUE),
    max_cost = max(Median_Delivery_Cost, na.rm = TRUE)
  )

print(cost_by_category)

# Visualizing the cost distribution by product type
ggplot(cost_by_category, aes(x = Applicable_To, y = avg_cost, fill = Applicable_To)) +
  geom_bar(stat = "identity") +
  labs(title = "Average Delivery Cost by Product Category",
       x = "Product Category",
       y = "Average Delivery Cost") +
  theme_minimal()

```





# Identify High-Cost Customer Segments

```{r}

```

